\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-10-26 document/XSGRB_sample.tex
Adding continuum estimate

\title{\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample}
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}

\titlerunning{XSGRB sample}

\author{
	J.~Selsing\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	P.~Goldoni\inst{14}, 
	D.~Malesani\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

\institute{Dark Cosmology Centre, Niels Bohr Institute, University of Copenhagen, Juliane Maries Vej 30, 2100 K\o benhavn \O, Denmark
\and European Southern Observatory, Alonso de C\'{o}rdova 3107, Vitacura, Casilla 19001, Santiago 19, Chile
\and Department of Particle Physics and Astrophysics, Faculty of Physics, Weizmann Institute of Science, Rehovot 76100, Israel
\and Th\"uringer Landessternwarte Tautenburg, Sternwarte 5, 07778 Tautenburg, Germany
\and Centre for Astrophysics and Cosmology, Science Institute, University of Iceland, Dunhagi 5, IS-107 Reykjavik, Iceland
\and Astronomical Institute Anton Pannekoek, University of Amsterdam, Science Park 904, NL-1098 XH Amsterdam, the Netherlands
\and INAF-Osservatorio Astronomico di Roma, Via Frascati 33, I-00040 Monteporzio Catone, Italy
\and GEPI-Observatoire de Paris, CNRS UMR 8111, Univ. Paris-Diderot, 5 Place Jules Janssen - 92190 Meudon, France
\and American River College, Physics and Astronomy Dpt., 4700 College Oak Drive, Sacramento, CA 95841, USA
\and INAF, Osservatorio Astronomico di Brera, Via E. Bianchi 46, I-23807 Merate, Italy
\and Instituto de Astrof\'{\i}sica de Andaluc\'{\i}a (IAA-CSIC), Glorieta de la Astronom\'{\i}a s/n, 18008, Granada, Spain
\and ASI-Science Data Centre, Via Galileo Galilei, I-00044 Frascati, Italy
\and Institute of Experimental and Applied Physics, Czech Technical University in Prague, Horska 3a/22, 128 00 Prague 2, Czech Republic
\and APC, Astroparticules et Cosmologie, Universite Paris Diderot, CNRS/
	IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cite, 10, Rue Alice Domon et L
	eonie Duquet, 75205 Paris Cedex 13, France
\and Max-Planck-Institut f\"{u}r extraterrestrische Physik, Giessenbachs
	tra\ss e, 85748 Garching, Germany
\and Universit\`a degli studi di Milano-Bicocca, Piazza della Scienza 3,
	20126, Milano, Italy
\and Pontificia Universidad Cat\'{o}lica de Chile, Departamento de Astro
	nom\'{\i}a y Astrof\'{\i}sica, Casilla 306, Santiago 22, Chile
\and Millennium Center for Supernova Science
\and Department of Physics and Astronomy, University ofLeicester, University Road, Leicester, LE1 7RH, UK
\and University of Ljubljana,Department of Physics,Faculty of Mathematics \& Physics,SI}

\abstract{
The {\it Swift} satellite allows us to use gamma-ray bursts (GRBs) to peer into 
the hearts of star forming galaxies through cosmic time. Our open
collaboration, representing most of the active ESO member researchers in this
field, seeks to build a public legacy sample of GRB X-shooter spectroscopy
while {\it Swift} continues to fly. We propose to continue our programme to target
all suitably observable GRB afterglows (up to 15 bursts per semester), with the
primary goal of producing a well-defined, homogeneous, statistically useful
sample. To date, our spectroscopy covers a redshift range from 0.059 to about
8, with more than 20 robust metallicity measurements from absorption lines (over 
the redshift range 1.7--5.9) and 4 secure detections of H$_2$ or CH molecular
absorption. Such information is extremely difficult to obtain by other means.
In terms of studying the spread and redshift evolution in gas-phase properties, the
sample is still limited by low-number statistics. 
}

@@

Gamma-Ray bursts (GRBs) are interesting as they provide the opportunity to
derive new insights into a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to the magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust extinction
curves, starburst galaxies, chemical and molecular abundances, escape of ionizing 
radiation, the ionization state of the intergalactic medium, intervening
absoption systems to standard candles in cosmology.  In that sense GRB research
is very multidisciplinary.

The {\it Swift} satellite \citep{Gehrels09}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.][]{Ricker04}.  A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. Low-resolution
spectroscopy is powerful for a number of reasons, first of all is the ability
to secure information for even very faint targets. This allows the measurements
of a number of important parameters such as redshifts and spectral slopes. For
a handful of very bright afterglows high-resolution (typically R$>$20000)
spectra were secured and for these events much more information about
conditions inside the host galaxies were extracted 
\citep[e.g.,][]{Fiore,Thoene07,Prochaska07,Vreejswijk07,CastroTirado2010}.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd
generation instruments at the ESO Very Large Telescope (VLT). It is a
high-throughput echelle spectrograph covering the full spectra range from the
UV atmospheric cutoff to the near-infrared (near-IR) K-band with three spectral
arms (UVB, VIS, NIR). It was designed very much with transient source follow-up
in mind as the fading luminosities of such sources makes in urgent to secure as
extensive coverage as possible in the shortest possible time. At the same time
the resolution was set to be in the range 4000--9000 in order to be able to get
a large useful spectra range between the many sky-background emission lines in
the red and near-IR spectral ranges. In this paper we present the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter
spectrograph to secure spectroscopic observations of
afterglows and host galaxies of GRBs detected by {\it Swift}. 
All data resulting from the survey are publicly availble in reduced form.

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy.  In Sec.t~\ref{results} we describe the results of the survey, i.e.
the efficiency of the follow-up effort and the characteristics of the observed
bursts. We also here assess the completeness of the finally realized sample.
Finally, we offer our conclusions in Sect.~\ref{conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and observations}\label{sample}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists from already, being it weather conditions, pointing
restrictions of the telescope or poorly localized bursts as reported by the
\textit{Swift}-telescope. To maximize the return of the follow-up campaign
we have chosen a few selection criteria that attempts to provide an unbiased
selection of bursts, while allowing for a high success-rate

\footnote{Note that in the P84 proposal the criteria have
been stated a bit differently, the visibility constraint being replaced by a
declination + Sun angle constraint. The above criteria are however those
defining the sample.}

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item Galactic $A_V \lesssim 0.5$ mag.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\item No bright closeby stars.

\end{enumerate}

A significant fraction of the bursts presented here \todo{Insert exact
number} have already had their hosts investigated in \citet{Kruhler2015}, for
which extractions of the hosts exist. The focus of the data presented here are
on the afterglows themselves and in the absence of a clear afterglow, the host.
We will not, however, investigate the hosts.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations. By restricting
the selection criteria to conditions local to Galaxy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

All the observations in this sample has been carried out using the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
two of the four Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal)
during the duration of this follow-up campaign, which covers the entire lifetime
of X-shooter. Observations have been carried out from the ESO period P84
through P98 under the following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079,
095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commisioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800
\AA~in a single exposure, by splitting the light in three separate
spectroscopic arms using dichroics.  This way each arm work as a separate
instrument, each functioning as its own echelle spectrograph.  The ultraviolet
blue(UVB) arm covering 3000 - 5500 \AA, the visual(VIS) arm covering 5500 -
10200 \AA, and the near-infrared()NIR) arm covering 10200 - 24800 \AA~ with the
possibility of applying a k-band blocking filter cutting the coverage of the
NIR arm at 10200 - 21000 \AA.
\\

\todo{ABBA}

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.
\\
\todo{ADC discussion}

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex}

@@

@@

@@

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/resolution_paper.eps}}
	\caption{Green datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric absorption lines in the VIS spectra, as a function of the FWHM of a Gaussian fit onto the trace in spatial direction at  792 nm. The lower horizontal axis is in units pixels, the top axis in arc seconds. The red datapoints show a subsample of NIR spectra.
		The grey line shows a linear fit to the VIS datapoints. }
	\label{fig:res}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
sample obtained is complete.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up timing and afterglow brightness}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A rapid follow-up is essential for the accurate designation of GRB the origin. The redshift determination of burst where the host is not easily

\begin{figure}
	\centerline{\includegraphics[width=8cm]{figures/timing.pdf}}
	\caption{Follow-up timing and afterglow magnitude at the start of observation. The points have been colored based on the redshift of the corresponding burst. }
	\label{fig:timing}
\end{figure}

>>>>

\title{XSGRB - The X-shooter GRB afterglow sample
\thanks{Based on observations collected at the European Southern Observatory, Paranal, 
	Chile, Program ID: 098.A-0055, 097.A-0036, 096.A-0079, 095.B-0811(B), 
	095.A-0045, 094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088,
	089.A-0067, 088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260}}

\titlerunning{The X-shooter sample of GRBs}

\input{authorlist.tex}

@@

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter up to the end of 2016. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% \todo{Preliminary} and tentatively confirm previous
results that larger optical darkness leads to increased X-ray absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

@@

GRBs are interesting as they provide the opportunity to derive new insights
into a very wide range of topics in astrophysics. In that sense GRB
research is very multidisciplinary. The {\it Swift} satellite \citep{Gehrels09}, 
which was launched in 2004, has made
it possible to harvest the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by
results from earlier missions.

A crucial aspect of the great success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of
the host galaxies of the GRBs. This work has involved a large community
of researchers both inside and outside the official {\it Swift} team. This fruitful
collaboration has been greatly facilitated by the open data access policy
of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra were
secured.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput 
echelle spectrograph
covering the full spectra range from the UV atmospheric cutoff to the near-infrared
(near-IR)
K-band with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources 
makes in urgent to secure as extensive coverage as possible in the shortest possible 
time. At the same time the resolution was set to be in the range 4000--9000 in order
to be able to get a large useful spectra range between the many sky-background
emission lines in the red and near-IR.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sample selection criteria and follow-up}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample selection criteria} \label{samplecrit}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs, without hampering the follow-up effort. Many natural follow-up
restrictions exists already, being it weather conditions, pointing restrictions
of the telescope or poorly localized bursts as reported by the alerting
telescope. To maximize the return of the follow-up campaign we have chosen a few
selection criteria that attempts to provide an unbiased selection of bursts,
while allowing for a high follow-up success rate. The importance of defining
unbiasing selection criteria has been highlighted previously
\citep{Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a}, when trying to
address the specific underlying distribution functions such as the redshift
distribution, host metallicity distribution, or brightness distribution. When
investigating a specific distribution function, a high degree of
\textit{completeness} is desired meaning that in terms of the parameter of
interest, the largest possible fraction of targets in the sample has
determinations of that parameter \citep{Perley2016b}.

We simultaneously want to minimize any biases against astrophysical conditions
while at the same time maximizing likelihood of observations allowing us to
obtain a higher degree of observing completeness. By restricting the selection
criteria to conditions local to Galaxy and therefore irrespective of intrinsic
GRB properties, the hope is that the sample collect represents the underlying
distribution of GRBs in a fair way while at the same time, additional unbiasing
cuts can be imposed that can improve the follow-up success. The selection
criteria used here is based on previous successes \cite{Jakobsson2006b,
	Fynbo2009, Hjorth2012}, which has shown to represent the underlying distribution
to a high degree. We test the degree of completeness in Sect.
\ref{completeness}.

The selection criteria are defined as follows:

\begin{enumerate}
	\item GRB triggered onboard by Swift.
	\item XRT started observing within 10 minutes since the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes at least 30 degrees above the horizon, with the Sun below -12.
	\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. The below criteria are however those
	defining the sample.}
	\item No bright closeby stars.
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

Because our ability to observe GRB afterglows is strongly correlated with the
rapidness and precision of the target positions delivered by the triggering
facilities, by selecting only bursts that have been triggered on board the
\textit{Swift} space telescope \cite{Gehrels2004}, by the Burst Alert Telescope
(BAT) \citep{Barthelmy2005}, we start out with a homogeneously selected sample
where burst characteristics are delivered immediately, allowing for an informed
follow-up strategy. Despite the complexity of the triggering mechanism on board
\textit{Swift} \citep{Band2006}, attempts at inferring properties of the
underlying cosmic GRB population based on the detection thresholds and
triggering algorithms have been made \citep{Lien2014, Graff2016} and thereby by
restricting the follow-up effort to bursts detected by \textit{Swift}, the
limitations of the parent sample is well studied.

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, we additionally require an X-ray position from the X-ray
Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} with 10 hours and to account for observing constraints on
\textit{Swift} that XRT began observations within 10 minutes. The additional
rapidness requirement of the XRT follow-up means that the number of detected
X-ray afterglows for BAT triggers is 100 percent \citep{Burrows2007}. The
\textit{Swift} constraints on the GRB follow-up means that the sample presented
here inherits the \textit{Swift} sample properties, except the additional XRT
constraint. Because the XRT completeness is very high\citep{Burrows2007}, this
should change the parent sample significantly. This issue is addressed again in
Sect. \ref{completeness}.

To allow for observability we additionally require that the GRB is visible
(duh!) from the telescope site at Cerro Paranal, Antofagasta, Región de
Antofagasta, Chile, for a least 1 hour allowing time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
seen from Earth and that GRB properties does not depend on position on the sky
\citep{Briggs1996}, this cut does not influence our ability to sample the
underlying GRB population. The same arguments applies to the requirement that
there are no nearby bright stars, which is additionally supplemented by the
requirement what the galactic dust content does not exert an extinction higher
than $A_V \lesssim 0.5$ mag based on the extinction maps by
\citet{Schlegel1998}. Because of  isotropy, these additional cuts should not
influence the optical properties of the bursts themselves, only our ability to
successfully secure observations that allows us to investigate the spectroscopic
properties of GRBs.

We address the succesfullness of these criteria and the actual follow-up effort
in representing the full \textit{Swift} sample ion Sect. \ref{completeness}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure and practicalities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Will be written by Daniele?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Observations} \label{obs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}}
	\caption{Mollweide projection in galactic coordinates of the full sky showing
		the positions on the sky of the bursts presented in this work. The empty
		circles are the positions of all the \textit{Swift} bursts detected until the
		end of 2016. In blue is shown the position of the bursts fulfilling the sample
		criteria specified in Sect. \ref{samplecrit}. The red circles are added around
		GRBs which have been followed up by X-shooter since the commissioning of the
		instrument. Blue points with red circles represent GRBs that enter our sample
		and have X-shooter spectroscopy. The different samples are compared in Sect.
		\ref{results}. The background shows the dust maps presented in
		\citet{Schlegel1998} where the sample criteria cut with $A_V \lesssim 0.5$ mag
		is removed.}
	\label{fig:skymap}
\end{figure*}

Observations in this sample has been carried out using the cross-dispersed
echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on two of the four
Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the duration
of this follow-up campaign, which covers the entire lifetime of X-shooter.
Observations have been carried out from the ESO period P84 through P98 under the
following programme IDs: 098.A-0055, 097.A-0036, 096.A-0079, 095.A-0045,
094.A-0134, 093.A-0069, 092.A-0124, 0091.C-0934, 090.A-0088, 089.A-0067,
088.A-0051, 087.A-0055, 086.A-0073, 085.A-0009 and 084.A-0260. A few additional
bursts, not a part of the \textit{statistical sample}, have been included from
092.D-0056(A) (PI: Rau), 092.D-0633(E) (PI: Greiner), and Levan under
095.B-0811(B) (PI: Levan). This sample represents \textit{all} GRBs that have
ever been followed up by X-shooter.

The first GRB followed up is GRB090313, observed 15th of March 2009, which was
while X-shooter was mounted on UT3 during the commissioning of the instrument.
The bursts observed during the commissioning period is not a part of our
statistical sample, but are nonetheless published as part of the bursts observed
by the X-shooter GRB team. The first burst observed after science verification
was completed when X-shooter was moved to UT2 is GRB091018 which thereby
constitute the first burst entering our statistical sample. For all bursts that
fulfill our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up has been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of this
effect is included in Sect. \ref{badbursts}.

X-shooter can cover the spectral wavelength region from 3000 \AA~to 24800 \AA~in
a single exposure, by splitting the light in three separate spectroscopic arms
using dichroics. This way each arm work as a separate instrument, each
functioning as its own echelle spectrograph. The ultraviolet blue(UVB) arm
covering 3000 - 5500 \AA, the visual(VIS) arm covering 5500 - 10200 \AA, and the
near-infrared()NIR) arm covering 10200 - 24800 \AA~ with the possibility of
applying a k-band blocking filter cutting the coverage of the NIR arm at 10200 -
21000 \AA. For all observations, a nod-and-shuffle observing scheme has been
employed, with a nodding throw of 5 \arcsec. The individual observations has
been carried out in a standard ABBA pattern, which allows flexibility in the sky
subtraction method for the different arms.

For the large majority of the bursts we have observed with a slit width of
1\farc0, 0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively, which
puts a lower limit on the delivered resolution of the spectra based on the
tabulated values of the delivered resolutions, which is 4350, 7450, and 5300 for
the UVB, VIS and NIR-arm respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}.
The slit width also sets the width of the atmospheric sky lines and determines
the amount of light lost due to the wavelength-dependent seeing PSF extending
outside the coverage of the slit, where the width of sky-lines is always set by
the slit width whereas both the delivered resolution and the slitloss changes
for the better as the seeing PSF drops below the slit width.
For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9
observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit,
meaning that for almost all observations a slitloss correction is required. We
describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements of velocity widths, a precise instrumental resolution
is required and this becomes better when the delivered seeing is better than the
slit width. We discuss this in Sect. \ref{resolution}.

Due to mechanical issues, the atmospheric dispersion corrector (ADC) were
disabled from 1st of August 2012. Only GRB~100728B was affected by the failing
ADC prior to disablement, resulting in a lower-than-nominal throughput. To avoid
chromatic slit losses due to atmospheric dispersion, all subsequent observations
have been carried out at parallactic angle. A consequence of this is that for
all observations following 1st of August 2012, the source moves across the
spatial direction of the slit as a function of wavelength. This effect has been
modeled in the extraction procedure, as described in \ref{extract}.

We list the overview of all the observations in Tab. \ref{tab:sample_overview}.
\input{tables/burst_overview.tex} and plot the positions off all the bursts on
the night sky in Fig. \ref{fig:skymap}. The central zone of avoidance due to
high galactic extinction cutoff is visible with an apparent isotropic
distribution of the bursts away from the galactic center except in the upper
right corner, where Earth casts a shadow on the sky as seen from the telescope.

A subset of the bursts presented here  have already had their hosts investigated
in \citet{Kruhler2015}, for which extractions of the hosts exist. The focus of
the data presented here are on the afterglows themselves and in the absence of a
clear afterglow, the host. We will not, however, investigate the hosts.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
	\caption{Telluric corrected, normalized spectrum of GRB121024A at $z = 2.300$,
	meant to illustrate the typical data quality. The acquisition magnitude is r =
	20, meaning it is in the brighter end of the sample presented here, but not the
	brightest. The spectrum is rich in absorption, including molecular absorption
	of $H_2$. The absorption trough visible at $\sim$ 4000 $\AA$ is due to \lya in
	the host. We have overplot the most prominent lines seen in GRB afterglows from
	\citet{Christensen2011a}. The continuum estimate is shown in dashed red.
	Additionally, three intervening systems are seen in this sightline  This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
	\label{fig:spectrum}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Continuum estimate} \label{continuum}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We additionally provide an estimate of the continuum for all the spectra
presented here. We have developed an algorithm that attempts to automatically
find the continuum placement along with the error on the continuum estimate. For
each arm,

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=8cm]{figures/resolution_paper.eps}}
	\caption{Green datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric absorption lines in the VIS spectra, as a function of the FWHM of a Gaussian fit onto the trace in spatial direction at  792 nm. The lower horizontal axis is in units pixels, the top axis in arc seconds. The red datapoints show a subsample of NIR spectra.
		The grey line shows a linear fit to the VIS datapoints. }
	\label{fig:res}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Science data products} \label{products}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB/} and additionally at , including the ones outside
the statistical sample. This release includes both prompt afterglow observations
as well as late time observations. An overview of all the spectra provided here
are given in Table. \ref{sample_overview}. For each burst, each individual
observation is provided in a separate reduction and in cases where observations
have been repeated, a combined spectrum is also provided.  No attempt has been
made to combine the arms, so for each observation, three spectra exists.

We have
adopted the following naming convention GRB*OB*.fits

In this section we describe the data format. Similar to what is done in \citet{Lopez2016}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results} \label{results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
